\documentclass{article}
\begin{document}
{\bf HOW DO YOU JUSTIFY YOUR EXISTENCE?}

Bidirectional situations are all around us. We synchronize bookmarks, keep
clones of our data on our mobile devices, edit collaboratively, use GUIs to
visualize and modify chunks of our data, etc. So designing tools that
support the creation and maintenance of these transformations is important.

Lots of nice things have been done (e.g. the asymmetric state-based lens
model and its instantiation in Boomerang), but there are many niches where
we might want more targeted support. For decentralized applications, neither
repository is canonical and you want a way to deal with that. For big data,
you want small descriptions of changes. Some data structures aren't most
naturally represented as strings, but as trees, relations, grids, graphs, or
other structures instead.

My thesis will be an exploration of these issues and what can be done to
deal with them; the major unfinished work that I will be proposing is
grid-structured data.

Outline the rest of the document now.
\begin{itemize}
    \item project proposal: bidirectional spreadsheet
        \begin{itemize}
            \item unidirectional spreadsheet: cells may have value or
                formula
            \item cells with formulas display the result of computing the
                formula
            \item when updating a cell, any formula depending on that cell
                is automatically updated
            \item to bidirectionalize: let each cell have \emph{both} value
                and formula
            \item update value independently from formula
            \item when updating a value: automatically compute new values
                for cells the formula depends on that would have that output
                as a result
        \end{itemize}
    \item why it's useful: big companies want to make predictions about
        their sales and do \emph{planning}: figuring out what local (e.g.
        store-specific or month-specific or product-specific) goals to set
        to achieve global goals
    \item TODO: examples, start with an easy one and then show what's hard
    \item why not just use a computer algebra system?
        \begin{itemize}
            \item spreadsheets are a familiar paradigm; also some research
                suggests that forming queries as a program up front (without
                being able to visualize intermediate outputs) requires
                significantly more skill than dealing with data directly
            \item want to deal with data that is not only numbers, e.g.
                product names, ID numbers, dates
            \item depending on which CAS you choose, guarantees of solution
                existence/uniqueness may be light/nonexistent -- perhaps
                some type system can help with this
            \item (pending literature search into just how exciting computer
                algebra gets) perhaps there are some operations we want --
                sorting, filtering, aggregations -- that are not readily
                available in CASs
        \end{itemize}
    \item why not just use a constraint satisfaction algorithm?
        \begin{itemize}
            \item once you've chosen an update, constraint satisfaction only
                allows information to flow one way; when there's more than
                one way to satisfy a constraint, one might like to use old
                values of the satisfied constraint to choose the new way to
                satisfy it
            \item (pending more literature search into how exciting
                constraint satisfaction gets) perhaps there is a story to
                tell about how to model the user guiding the constraint
                satisfaction process when there's many ways to satisfy a
                constraint
        \end{itemize}
    \item TODO: planned contributions
    \item TODO: also, timeline, failure modes, related work, etc.
\end{itemize}
\end{document}
