\documentclass{eceasst}
\usepackage{paper}
\begin{document}
\title{Edit languages for information trees}
\author{Martin Hofmann and Benjamin C. Pierce and Daniel Wagner}
\include{frontmatter_tree}
\institute{Ludwig-Maximilians-Universit\"at and University of Pennsylvania}
\keywords{Bidirectional programming, lens, edit, information tree}

\abstract{We consider a simple set of edit operations for unordered,
  edge-labeled trees, called {\em information trees} by Dal Zilio {\em et
  al}~\cite{DalzilioS:POPL04}.  We define {\em tree languages} using the
  {\em sheaves automata} from \cite{Foster:FTL} which in turn are based on
  \cite{DalzilioS:POPL04} and provide an algorithm for deciding whether a
  complex edit preserves membership in a tree language. This allows us to
  view sheaves automata and subsets of tree edits as {\em edit languages} in
  the sense of~\cite{HofmannPierceWagner12}. They can then be used to
  instantiate the framework of {\em edit lenses} between such languages and
  model concrete examples such as synchronisation between different file
  systems or address directories.}
\maketitle

\section{Introduction}

\iflater
\reviewcomment{The paper would be more appealing if strengthened by
  examples. Early on. Specifically, the example application mentioned in the
  last paragraph of Section 4 could be concretized, elaborated, and
  presented first as motivation, then in its actual realisation, up front
  and then as running example along with the formal concepts and
  constructions.}

\reviewcomment{Also concerning applicability, I have a general question
  (also of theoretical interest, I think) about the chosen setup: What about
  more powerful edits? Currently, no recursion or iteration is
  supported. Probably it is not even desirable in full generality, but what
  about iteration at least during navigation? The at(n,e)-edit is only
  one-step navigation. It would not allow to capture something like the
  transitive axes of XPath. Can you capture those, by extending the edit
  language (but maintaining the decidable type-checking)?  }
\fi

Semantic models of bidirectional transformations are generally presented as
transformations between the {\em states} of replicas.  For example, the
familiar framework of asymmetric lenses defines a lens between replicas of
types $A$ and $B$ as a pair of a {\em get} function from $A$ to $B$ and a
{\em put} function from $A \times B$ to $A$.
%
An implementation directly based on this semantics would pass to the {\em
  put} function the entire states of the original $A$ and updated $B$
replicas.

Though pleasingly simple, this treatment falls short of telling a full story
in at least two important ways.
First, it does not explain how the lens should {\em align} the parts of the
old $A$ and the new $B$ so that the parts of $A$ that are ``hidden'' in the
$B$ view retain their positions in the result.  For example, if $A$ and $B$
are both lists of people where each element of $A$ includes a name, address,
and email while the corresponding elements of $B$ give only a name and
address, the user will reasonably expect that inserting a new element at the
head of the $B$ replica will lead to an updated $A$ replica where each
existing name keeps its associated email.  And second, the simple
``classical'' form of asymmetric lenses fails to capture the reasonable
expectation that a small change to the $B$ replica can be transformed into a
change to the $A$ replica using time and space proportional to the size of
the change, not to the sizes of the replicas.  \finishlater{This second point
  sounds a bit bogus, since the system we have in mind is going to have to
  walk over the whole replicas to compute the edit from the original and
  modified structures in the first place. Martin: but then once the edit has been computed one only needs to send a small bit of info}

One way to address at least the first concern is to enrich the basic
structure of a state-based lens with a new input to the {\em put} function,
a data structure that explicitly represents the {\em alignment} between the
original and updated $B$ replicas; this idea forms the basis for {\em
  dictionary lenses}~\cite{Boomerang07}, {\em matching
  lenses}~\cite{Matching10}, and {\em symmetric}~\cite{Diskin-Delta11} and
{\em asymmetric delta lenses}~\cite{diskin2011asymmetric} (based on
\cite{Stevens07}).\finishlater{Need to cite Perdita's relevant papers
  too!}  Another approach is to annotate the $B$ structures themselves with
change information~\cite{HuModels07, Hidaka10}.  However, all these
approaches still involve whole replica states, either as explicit inputs and
outputs of the {\em put} function or implicitly as part of the type to which
a delta belongs.  Thus, it is not clear whether these models can be
implemented in such a way that small changes to a replica are propagated in
time and space proportional to the size of the change.

\finishlater{Maybe we also need
  to cite some of these?  \cite{Meertens98, HuEditor08, MuAlgebraic2004,
    xiong09}}

In earlier work~\cite{HofmannPierceWagner12}, we proposed going a step further and defining
lenses that work exclusively with edits.  We defined a semantic model called
{\em edit lenses} in which the sets of source and target replicas $A$ and
$B$ are enriched with monoids of {\em edits} and a lens's {\em get} and {\em
  put} functions map edits to edits.  This work was carried out in an
abstract algebraic setting where the actual data structures being
transformed and the exact shapes of edits to them were left unspecified; in
this setting, we showed how a number of familiar constructions on
lenses---products, sums, etc.---could be carried out.

The present paper takes a first step
toward instantiating this abstract semantic
model with concrete data structures and a concrete notion of edits.  The
data model we choose is a very common and expressive one: unordered,
edge-labeled trees---called {\em information trees} by Dal Zilio {\em et
  al}~\cite{DalzilioS:POPL04}. These can encode a variety of data formats,
including XML-style trees, their original application.
Such trees can also be used to represent graphs
\cite{UnQL96,Hidaka10} by unrolling up to bisimulation. This paper thus
makes a first step towards general edit languages for trees.  Our edit
operations include, in particular, insertion and deletion of subtrees and
renaming of edges; we show how these give rise to edit languages on tree
languages and can thus be used to instantiate the framework of {edit
  lenses} so as to yield bidirectional synchronization between information
trees.

Our main technical result is that weakest preconditions of our edits
can be effectively computed for sets of information trees specified by
sheaves automata. This allows for effective ``type checking'' of edits
and thus permits automatic checks that an edit language for trees
presented as sequences of atomic edits does indeed preserve acceptance by a
given sheaves automaton.

The present work is thus a first step; we hope that the introduction
of tree automata into the world of editing and synchronization will
also lead to high-level support for constructing edit lenses
themselves and for checking their soundness, but this remains future
work.

\input{tree}

\section{Conclusion}
We have defined a simple set of edits for information trees comprising insertions, deletions, relocations, and renamings of subtrees. Our main technical result states that tree languages defined by sheaves automata \cite{DalzilioS:POPL04} are effectively closed under weakest preconditions for these edits and that therefore, typechecking of edits against tree types defined by sheaves automata \cite{Foster:FTL} is algorithmically tractable.

We see this result as a first step towards an automata-based high-level formalism for tree synchronisation; in particular we would like to investigate to what extent complements and consistency relations can be defined by automata and how the tree types from \cite{Foster:FTL} and the associated term formers can be lifted to edit lenses. More speculative goals include the automatic discovery of tree edits (``tree diffing'') and the extension to graphs.

\bibliographystyle{eceasst}
\bibliography{tree,bcp,harmony}

\end{document}
