\iffull
\documentclass[prodmode,acmjacm]{acmsmall}
\else
\documentclass{sigplanconf}
\fi

\usepackage{paper}

% \title{Delta Lenses: Operation-based Bidirectional Programming}
% \title{Two Algebras of Lenses}
%% The Category of Change
%% Quick-Change Artists
% \title{Polishing Lenses From Raw Materials}
% \title{Elements of Lenses}
% \title{Elements of Symmetric Lenses}
\iffull
\acmVolume{V}
\acmNumber{N}
\acmArticle{A}
\acmYear{YYYY}
\acmMonth{0}
\markboth{Martin Hofmann et al}{Symmetric Lenses}
\title{Symmetric Lenses}
\author{MARTIN HOFMANN \affil{Ludwig-Maximilians-Universit\"at} BENJAMIN
PIERCE and DANIEL WAGNER \affil{University of Pennsylvania}}
\else
\begin{document}
\title{Symmetric Lenses}
\authorinfo
    {Martin Hofmann}
    {Ludwig-Maximilians-Universit\"at \\ M\"unchen, Germany}
    {}
\authorinfo
    {Benjamin Pierce}
    {University of Pennsylvania \\ Pennsylvania, USA}
    {}
\authorinfo
    {Daniel Wagner}
    {University of Pennsylvania \\ Pennsylvania, USA}
    {}
\conferenceinfo{POPL'11,} {January 26--28, 2011, Austin, Texas, USA.}
\CopyrightYear{2011}
\copyrightdata{978-1-4503-0490-0/11/01}
\maketitle
\fi

\iftext
\begin{abstract}
{\em Lenses}---bidirectional transformations between pairs of connected
structures---have been extensively studied and are beginning to find their
way into industrial practice. However, some aspects of their foundations
remain poorly understood.
%
In particular, most previous work has focused on the special case of {\em
  asymmetric lenses}, where one of the structures is taken as primary and
the other is thought of as a projection, or view.  A few studies have
considered symmetric variants, where each structure contains
information not present in the other, but these all 
lack the basic operation of {\em composition}.  Moreover,
while many domain-specific languages based on lenses have been designed,
lenses have not been thoroughly explored from an algebraic
perspective.

We offer two contributions to the theory of lenses.  
%
First, we present a new symmetric formulation, based on {\em complements},
an old idea from the database literature.  This formulation generalizes the
familiar structure of asymmetric lenses, and it admits a good notion of
composition.
%% The cost of
%% this generality is that we often need to reason about these lenses modulo a
%% behavioral equivalence.  
Second, we explore the algebraic structure of the space of
symmetric lenses.  We present generalizations of a number of known
constructions on asymmetric lenses and settle some longstanding questions
about their properties---in particular, we prove the existence of (symmetric
monoidal) tensor products and sums and the {\em non-}existence of full
categorical products and sums in the category of symmetric lenses.  We then
show how the methods of universal 
algebra can be applied to build {\em iterator lenses} for structured data
such as lists and trees, yielding lenses for operations like
mapping, filtering, and concatenation from first principles.  Finally, we
investigate an even more general technique for constructing mapping
combinators, based on the theory of {\em containers}.
\end{abstract}
\fi

% Included below for the syntax, but not filled in (yet)
\category{D.3.2}{Programming Languages}{Language
Classifications}[Specialized application languages]
\terms{Design, Languages, Theory}
\keywords{algebra, category theory, lens, view-update}

\iffull
\acmformat{Martin Hofmann et al 2011. Symmetric Lenses.}
\begin{document}
\begin{bottomstuff}
Supported by the National Science Foundation under grants IIS-0534592,
\emph{Linguistic Foundations for XML View Update}, and CCF-1017212, {\em Algebraic Foundations for Collaborative Data Sharing}.
\end{bottomstuff}
\maketitle
\fi
\section{Introduction}

The electronic world is rife with partially synchronized data---replicated
structures that are not identical but share some common parts, and
where the shared parts need to be kept up to date as the structures change.
Examples include databases and materialized views, in-memory and on-disk
representations of heap structures, connected components of user interfaces,
and models representing different aspects of the same software system.

In current practice, the propagation of changes between connected structures
is mostly handled by {\em ad hoc} methods. Given a pair of structures $X$
and $Y$, we write one
transformation that maps changes to $X$ into changes to $Y$ and a separate
transformation that maps $Y$ changes to $X$ changes.  However, when the
structures 
involved are complex, managing such pairs of transformations manually can be a
maintenance nightmare.

This has led to a burgeoning interest in {\em bidirectional programming
  languages}, in which every expression denotes a related pair of
transformations.  A great variety of bidirectional languages have been
proposed (see~\cite{FosterThesis,DBLP:conf/icmt/CzarneckiFHLST09} for recent
  surveys), and these ideas are 
beginning to see commercial application, e.g., in RedHat's system
administration tool, Augeas~\cite{Augeas}.

One particularly well-studied
class of bidirectional programming languages is the framework of {\em
  lenses}~\cite{Focal2005-shortcite}.
%
Prior work on lenses and lens-like structures has mostly been carried out in
specific domains---designing combinators for lenses that work over
strings~%
\iffull
  \cite{Boomerang07,QuotientLenses08,updatable-security-views,Matching10}, 
\else
  \cite{Boomerang07,QuotientLenses08,Matching10}, 
\fi
trees~\cite{Focal2005-shortcite,Hu04,MuAlgebraic2004,HuXQuery08}, 
relations~\cite{BohannonPierceVaughan}, graphs~\cite{Hidaka10}, or software
models~%
\iffull
\cite{Stevens07,HuModels07,stevens2008lbm,%
DBLP:conf/models/Diskin08,diskin2008algebraic,%
diskin2009model,hidaka2009compositional,hidaka2009towards}.
\else
  \cite{Stevens07,HuModels07, stevens2008lbm, 
    DBLP:conf/models/Diskin08, 
    diskin2009model, hidaka2009compositional}.
\fi
%
By contrast, our aim in this paper is to advance the {\em foundations} of
lenses in two significant respects.

%% \finish{Another citation that we might include (but it's not directly
%%     relevant): \cite{voigtlander2009bidirectionalization} }

First, we show that lenses can be generalized from their usual asymmetric
presentation---where one of the structures is always a ``view'' of the
other---to a 
fully {\em symmetric} version where each of the two structures may 
contain information that is not present in the other (\S\ref{symm}).
This generalization is significantly more expressive than any previously
known: although symmetric variants of lenses have been
studied~\cite{Meertens98,stevens2008tat,DBLP:conf/models/Diskin08}, they
all lack a notion of sequential composition of lenses, a significant
technical and practical limitation (see \S\ref{sec:relwork}).
%
As we will see, the extra structure that we need to support composition is
nontrivial; in particular, constructions involving
symmetric lenses need to be proved correct modulo a notion of {\em behavioral
  equivalence} (\S\ref{equiv}).

Second, we undertake a systematic investigation of the {\em
  algebraic structure} of the space of lenses, using the concepts of
elementary category theory as guiding and organizing principles.  (Our
presentation is self contained, but some prior familiarity
with basic concepts of category theory will be helpful.\iffull \else{} Most
  proofs 
are omitted for brevity; they can be found in a long
version of the paper, available from the second author's web page.\fi)

We begin this algebraic investigation with some simple generic constructions
on symmetric 
lenses: composition, dualization, terminal lenses, simple bijections, etc.
(\S\ref{basic}). We then settle some basic questions about products
and sums (\S\ref{prod} and \S\ref{sumlist}). In particular, it was
previously known 
that asymmetric lenses admit constructions intuitively corresponding to
pairing and projection~\cite{Boomerang07} and another construction that is
intuitively like a sum~\cite{Focal2005-shortcite}. However, these
constructions were not very well understood; in particular, it was not known
whether the pairing and projection operations formed a full categorical
product, while the {injection arrows} from $X$ to $X + Y$ and from $Y$ to
$X + Y$ were not definable at all in the asymmetric setting. We prove that
the category of symmetric lenses does \emph{not} have full categorical
products or sums, but does have ``symmetric monoidal'' structures
with many of the useful properties of products and sums.

Next, we consider how to build lenses over more complex data
structures such as lists and trees (\S\ref{iter}).  We first observe
that the standard construction of algebraic datatypes can be lifted
straightforwardly from the category of sets to the category of lenses.  For
example, from the definition of lists as the least solution of the equation
$L(X) \simeq \Unit + X \times L(X)$ we obtain a lens connecting the set
$L(X)$ with the set $\Unit + X \times L(X)$; the two directions of this
lens correspond to the \emph{unfold} and \emph{fold} operations on lists.
Moreover, the familiar notion of {\em initial algebra} also generalizes to
lenses, giving us powerful iterators that allow for a modular definition of
many symmetric lenses on lists and trees---e.g., mapping a symmetric lens
over a list, filtering, reversing, concatenating, and translating between
lists and trees.
%  
%% Furthermore, back-to-back composition of iterators in the style of {\em
%%   hylomorphisms} provides a pattern for (terminating) recursive definition
%% of symmetric lenses between arbitrary (not necessarily inductive)
%%   sets.\bcp{Not sure where that last comment belongs.  Or whether, perhaps,
%%   it's too detailed for the intro and is better just left to explain when we
%%   get to it.  Since we don't have any good examples at the moment, the whole
%%   point about hylomorphisms seems a bit of a digression.  It also weakens
%%   the contribution that we can claim in the next paragraph.}
  
Finally, we briefly investigate an even more general technique for
constructing ``mapping lenses'' that apply the action of a given sublens to
all the elements of some data structure (\S\ref{contain}).  This
technique applies not only to algebraic data structures but to an arbitrary
{\em container} in the sense of Abbot, Altenkirch, and
Ghani~\cite{1195941}.  This extends the variety of list and tree mapping
combinators that we can construct from first principles to include
non-inductive datatypes such as labeled dags and graphs.  
\iflater
\bcp{This
  still seems weak.  Can we make it sound more substantial / useful?}
\fi

%% Finally---our most technically substantial contribution---we consider lenses
%% for recursive datatypes such as lists and trees.  \S\ref{sumlist}
%% gives direct definitions of two variants of a ``list mapping'' constructor
%% (corresponding to the two variants of sum).  \S\ref{iter}
%% and~\ref{contain} develop two different ways to obtain the same lens as an
%% instance of a much more general construction.  In \S\ref{iter} we use
%% the methods of universal algebra \finish{... fill this in ...}.  In
%% \S\ref{contain} we use Joyal's notion of {\em
%%   containers}~\cite{JoyalContainers} \finish{... fill this in ...}.
%% \PENDING{This paragraph ought to be the most compelling in the whole intro
%%   because it describes our (in some sense) main results, but it's definitely
%%   not!}

We carry out these investigations in the richer space of symmetric lenses,
but many of the results and techniques should also apply to the special case of
asymmetric lenses.  Indeed, we can show (\S\ref{asymm}) that
asymmetric lenses form a subcategory
of symmetric ones in a natural way: every
asymmetric lens can be embedded in a symmetric lens, and many of the
algebraic operators on symmetric lenses specialize to known constructions on
asymmetric lenses.  Conversely, a symmetric lens can be factored into a
``back to back'' assembly of two asymmetric ones.

\S\ref{sec:relwork} and \S\ref{sec:future} discuss related and future
work. 

\input{complement}

\iftext
\section{Related Work}\label{sec:relwork}

There is a large literature on lenses and related approaches to
propagating updates between connected structures.  We discuss only the most
closely related work here; good general surveys of the area can be found
in~\cite{FosterThesis,DBLP:conf/icmt/CzarneckiFHLST09}.  Connections to the
literature on {\em view update} in databases are surveyed
in~\cite{Focal2005-shortcite}. \iffull A short version of this paper is available
in~\cite{HofmannPierceWagner10}.\fi

The first symmetric approach to update propagation was proposed by
Meer\-tens~\cite{Meertens98} and followed up \iffull in the context of
model-driven design \fi by Stevens~\cite{Stevens07},
Diskin~\cite{DBLP:conf/models/Diskin08}, and Xiong, et
al~\cite{xiong2009supporting}.
%
Meertens suggests modeling synchronization between two sets
$X$ and $Y$ by a {\em consistency relation} $R\subseteq
X\times Y$ and two {\em consistency maintainers}
$\triangleleft: X\times Y\rightarrow X$ and $\triangleright: X\times Y
\rightarrow Y$ such that $(x\triangleleft y) \relR y$ and
$x \relR (x\triangleright y)$ always hold, and such that $x \relR y$ implies
$x \triangleleft y = x$ and $x \triangleright y = y$.

The main advantage of symmetric lenses over consistency maintainers is
their closure under composition. Indeed, all of the aforementioned
authors note that, in general, consistency maintainers do not compose
and view this as a drawback.
%
Suppose that we have relations $R\subseteq X\times Y$ and
$R'\subseteq Y\times Z$ maintained by $\triangleright,\triangleleft$
and $\triangleright', \triangleleft'$, resp. If we want to construct a
maintainer for the composition $R;R'$, we face the problem that, given
$x\in X$ and $z\in Z$, there is no canonical way of coming up with a
$y\in Y$ that will allow us to use either of the existing maintainer
functions. Concretely, Meertens gives the following counterexample.
Let $X$ be the set of nonempty context free grammars over some alphabet, and let
$Y$ be the set of words over that same alphabet. Let $R\subseteq
X\times Y$ be given by $G \relR x\iff x\in L(G)$. It is easy to define
computable maintainer functions making this relation a constraint
maintainer. Composing this relation with its opposite yields an 
undecidable relation (namely, whether the intersection of two context-free
grammars is nonempty), so there cannot be computable maintainer functions.

We can transform any constraint maintainer into a  symmetric lens as
follows: take the relation $R$ itself (viewed as a set of pairs) as
the complement, and define $\putl(x',(x,y))=(x'\triangleright
y,(x',x'\triangleright y))$ and similarly 
for $\putr$. If we compose such a symmetric lens with its opposite
we obtain $R\times R\op$ as the complement and, for example,
$\putr(x',((x_1,y_1),(y_2,x_2))) = 
(x_2\triangleleft(x'\triangleright y_1), ((x',x'\triangleright
y_1),(x'\triangleright y_1,x_2\triangleleft(x'\triangleright y_1))))$. 
%
For Meertens' counterexample, we would have complements of the form
$((G_1,w_1),(w_2,G_2))$, with $w_1\in L(G_1)$
and $w_2\in L(G_2)$; ``$\putr$''-ing a new grammar
$G_1'$ through the composed lens yields the complement
$((G_1',w_1'),(w_1',G_2'))$, where $w_1'$ is $w_1$ if $w_1\in L(G_1)$ and
some default otherwise, and where $G_2'=G_2$ if $w_1'\in L(G_2)$ and
$S{\rightarrow}w_1'$ (where $S$ is the start state) otherwise. We observe
that there is a property of lenses analogous to Meertens' requirement that
$x \relR y$ implies $x \triangleleft y = x$. This property is not
necessarily preserved by composition, and in particular the lens described
above for synchronizing languages does not have it.
%
Meertens recommends using a {\em chain} of consistency maintainers in such a
situation to achieve a similar effect; however, the properties of such
chains have not been explored.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For asymmetric lenses, a number of alternative \iffull choices of behavioral
\fi
laws 
have been explored.  Some of these are \iffull strictly \fi weaker than ours; for
example, a number of papers from a community of researchers based in Tokyo
replace the \rn{PutGet} law with a somewhat looser \rn{PutGetPut} law,
permitting a broader range of useful behaviors for lenses that duplicate
information.  It would be interesting to see what kind of categorical
structures arise from these choices.  The proposal by Matsuda et
al.~\cite{matsuda2007btb} is particularly interesting because it also employs
the idea of complements.  Conversely, stronger laws can be imagined, such as
the \rn{PutPut} law discussed by Foster et
al.~\cite{Focal2005-shortcite}\iffull{} and the more refined variants
in~\cite{updatable-security-views}\fi\iflater\finish{Say something more
  about this?}\fi.

A different foundation for defining lenses by recursion was explored by
Foster et al.~\cite{Focal2005-shortcite}, using standard tools from domain
theory to define monotonicity and continuity for lens combinators
parametrized on other lenses.  The main drawback of this approach is that
the required (manual) proofs that such recursive lenses are total tend to be
somewhat intricate.  By contrast, we expect that our initial-algebra
approach can be equipped with automatic proofs of totality (that is, choices
of the weight function $w$) in many cases of interest.

\ifdelta
\finish{
Updates: (note that the point in our paper is not update as such but rather
how to marry it with lenses)  \finish{BCP will look for a good canonical
  reference for updates}
\begin{itemize}
\item lens-like things that have used deltas~\cite{Hu04,Meertens98,MuAlgebraic2004,HuModels07}
\item old update transformation
\item old view update stuff
\item new Waterloo stuff 
\begin{itemize}
\item flexibility, not efficiency
\item no proposal for syntax---composition is the only operator they study
at all
\end{itemize}
\end{itemize}
}
\fi

\section{Conclusions and Future Work}\label{sec:future}

We have proposed the first notion of symmetric bidirectional transformations
that supports composition.
%
Composability opens up the study of symmetric bidirectional
transformations from a category-theoretic perspective. We have 
explored the category of symmetric lenses, which is self-dual and has
the category of bijections and that of asymmetric lenses each as
full subcategories. We have surveyed the structure of 
this category
and found it to admit tensor product structures that are the
Cartesian product and disjoint union on objects.  We have also
investigated datatypes both inductively and as ``containers'' and found
the category of symmetric lenses to support powerful mapping and
folding constructs.

\paragraph*{Syntax}
Although we have focused here on semantic and algebraic foundations, many of
our constructions have a straightforward syntactic realization.  In
particular, it is easy to give a string-transformation interpretation to all
the constructions in \S\ref{basic} through \S\ref{sumlist} (including lenses
over lists); these could easily be used to build a symmetric version of
Boomerang~\cite{Boomerang07}. 
\iffull

\fi
More interesting would be to eliminate Boomerang's built-in lists and
instead obtain 
lenses over lists and other structures
(mapping, reversing, flattening
of lists, transforming trees into lists) solely
by using the combinators derived from the category-theoretic structure
we have exhibited. 
%
To accomplish this, two further fine points need to be considered. First, we would want 
an automatic way for discovering weight functions for iterators. We
believe that a straightforward termination analysis based on unfolding
(similar to the one built into Coq) could help, but the details remain to
be checked. And second, we must invent a formal syntax for programming
with containers. Surprisingly, the existing literature does not seem
to contain such a proposal.

More speculatively, it is a well-known folklore result that symmetric
monoidal categories 
are in 1-1 correspondence with wiring
diagrams\iffull~\cite{selinger2011survey}\fi\ and with first-order
linear lambda calculus. We would like to exploit this correspondence to design a
lambda-calculus-like syntax for symmetric lenses and perhaps also a
diagrammatic language. The linear lambda calculus has judgments of
the form $x_1{:}A_1,\dots,x_n{:}A_n\vdash t:A_0$, where $A_0,\dots,A_n$
are sets or possibly syntactic type expressions and where $t$ is a
linear term made up from basic lenses, lens combinators, and the
variables $x_1,\dots,x_n$. This could be taken as denoting a symmetric lens
$A_1\otimes\dots\otimes A_n\lens A_0$. For example, here is such a term for
the lens $\mathit{concat}'$ from \S\ref{concatprime}:  
\[\begin{array}{@{}l}
z{:}\Unit \oplus A \otimes A\LIST\otimes A\LIST\vdash 
\begin{array}[t]{@{}l}
\textit{match}\ z\ \textit{with} \\
\quad \mid\mlinl\unit \mapsto \const_{\NIL}\op\\
\quad\mid\mlinr(a,al,ar)\mapsto \mathit{concat}(a\CONS al,ar)
\end{array}
\end{array}
\]
The interpretation of such a term in the category of lenses then takes
care of the appropriate insertion of bijective lenses for regrouping
and swapping tensor products.

\paragraph*{Complements as States}
One benefit of treating complements explicitly is that it opens the way
to a {\em stateful} presentation of lenses. \iffull\else An upcoming workshop paper will
discuss our work on a Haskell implementation of symmetric lenses in this
vein. \fi The idea is that the complement
of a lens can be thought of as its local storage---the part of the heap that
belongs to it.  An obvious next step is that, instead of the lens components
taking the local storage as an argument and returning an updated version as
a result, they can just hang onto it themselves, internally, in mutable
variables.  The types of the \PUT{} operations then become just 
%
\iffull
\[
\begin{array}{r@{\ \;}c@{\ \;}l}
\ell.\PUTR &\in& A \arrow B \\
\ell.\PUTL &\in& B \arrow A
\end{array}
\]
\else
$\ell.\PUTR \in A \arrow B$ and 
$\ell.\PUTL \in B \arrow A$,
\fi
%
where the $\arrow$ is now a ``programming language function type,'' with the
usual implicit treatment of the heap.
%
%% The pleasant thing about this rearrangement is that it changes nothing in
%% the theory.  We can still model lenses in the same way, in a
%% ``mathematical,'' complement-passing style, but we can implement them with
%% mutable state, which will probably be more efficient, since it
This avoids
destructing the given $C$ each time we propagate an update and rebuilding a
new $C$ to yield as a result, improving the efficiency of the
implementation.  An additional improvement comes from 
the next potential extension.

\paragraph*{Alignment and Delta Lenses}
As we mentioned in \S\ref{firstalign}, dealing correctly with
alignment of structured information is crucial in practice.  This issue has
been extensively explored in the context of asymmetric lenses, and it seems
it should be possible to adapt existing ideas such as {\em dictionary
  lenses}~\cite{Boomerang07} and {\em matching lenses}~\cite{Matching10} to
symmetric lenses.  An even better approach might be to change the
fundamental nature of lenses so that, instead of working directly with
entire {\em structures}, they work with {\em deltas}---descriptions of
changes to the structures.  These deltas can arise from simple positional
judgments, as in this paper, from diff-like heuristics, from cues within
the data itself, or perhaps even from user interaction---the lens itself
doesn't need to know anything about this.  

Many of our basic constructions can be adapted to deltas by taking the
domain and codomain of a lens to be monoids (of edit operations) instead of
sets, and then, for each lens construction, defining an appropriate edit
monoid from the monoids of its components.  For example, an edit for a pair
lens is a pair of edits for the left- and right-hand sides of the pair.
However, more thought is required to make this scheme really work: applying
this idea naively leads to insufficiently expressive edit languages for
structures like lists.  In
particular, we would like to see insertion and deletion as edit
operations on lists (and rotations and the like for trees, etc.).
Currently, we believe that containers are a promising framework
for this endeavour.

\paragraph*{Algebraic Structure}
We have shown that the category of symmetric lenses does not have products
or sums.  One reviewer suggests that it may be interesting to explore
placing a partial order on the the homsets of our category (the sets of
arrows between pairs of objects). Perhaps this extra structure would allow
us to define variants of products and sums in which the required equations
do not hold as equalities, but as inequalities.

\fi % text
\iftext
\iffull
\begin{acks}
We are grateful to Nate Foster for productive discussions of many points,
especially about the category of lenses, and to both Nate and Alexandre
Pilkiewicz for comments on drafts of the paper.  The anonymous POPL
reviewers and JACM reviewers also made numerous helpful suggestions.  \finish{Anybody else we
  should cite?}
\end{acks}
\else
\SMALLSECTIONHEADER{Acknowledgments} We are grateful to Nate Foster for
productive discussions of many points, especially about the category of
lenses, and to both Nate and Alexandre Pilkiewicz for helpful comments on
drafts of the paper.  Our work is supported by the National Science
Foundation under grant IIS-0534592 \emph{Linguistic Foundations for XML View
  Update}.
\fi % full
\fi % text


\let\section\OLDsection

\iffull
\bibliographystyle{acmsmall}
\else
\bibliographystyle{plainnat}
\fi
\bibliography{bcp,harmony,complement,dissertation}

\end{document}


typeset this proof and corollaries.

reverse concatenate

typesetting trees

related work

rosebrugh paper


write intro to iterators section, say that we only look at retentive variant

finish conclusions





Comments that shouldn't necessarily result in a change to the paper:

It would be better to emphasize the importance of design of "suitable" 
complements. This is because one can design a trivial complement to 
make any non-injective function to be injective and such a bijective 
function can be executed in two directions.

Rationale: Surely being able to bidirectionalize any function would be a
huge win.

Existing work requires "complement" to be constant, but this paper 
allows the change on complements. What is the relationship between 
them?

Rationale: It is not clear that the most closely related work involves
constant complement. e.g. the existence of the fiat lens in the asymmetric
world seems to say to me that they don't do constant complement.

Is lens equivalence decidable?

Rationale: Maybe not, but a decision procedure doesn't seem necessary.

6.1: While it says (in the paragraph before the definition) that the 
tensor sum lens can switch between the two lenses as needed, the 
definition does not show this. I think it is the union lens (defined 
later) that plays the role of this switching. 

Rationale: Well, it does switch between the two lenses as needed to
accomodate the input -- it uses k when given an \mlinl and uses \ell when
given an \mlinr.

Would it be clear to specialize it under the context of lists and trees?

Rationale: This is another possible design decision.

What kind of updates is allowed for fold, hylo, ...?

Rationale: There's no concept of updates, but any new value on either side
can be transported -- we have no partiality. Although perhaps it is worth
stressing this fact even more in the paper somehow...?

I found myself thinking a lot about alignment. I'm not sure I like 
it. My instinct is that it would be much better to work with delta 
lenses.

Rationale: We totally agree! Stay tuned.

I think it's worthwhile to include the alternative formulation of
"observational" equivalence here, and the proof it conicices with def 3.2.
This is, in my view, more important than listing all the small propositions
later.

Rationale: space





Comments that haven't been handled, either by a change to the paper, a
\finish mark, or mark that they don't require changes:

- I am not so sure about the powerful/advantage of using symmetric 
  lenses in practice. Although symmetric lenses can deal with the 
  example in Figure 1 while asymmetric lenses cannot do, the work by 
  Yingfei et. al. (ICMT 2009) shows that asymmetric lenses, if being 
  combined with some effective "diff" and "merge" operators, can deal 
  synchronization of two data each contains information that is not in 
  the other.

While theoretically interesting, the updatability of the designed 
symmetric lenses are not very clear. For instance, although iteration 
can be used to define various functions on data structures, it does 
not seems to allow flexible change on shapes of the data.

dmwit says: I'm not sure I understand this comment. Is this attacking the
maintenance of a lens-based program, in which the representation of the data
on the two sides of the lens is changing as the program evolves to meet new
needs, or is it attacking the way our presentation of lenses transports a
new piece of data with the same type but a different shape?

The composition of two symmetric lenses is not "local"; it uses a 
"global" complement (to gather the two components).

The discussion of the tensor made me wonder whether it is possible to define
some order structure on the homsets so that the failing equations are
expressible as inclusions?
